BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-01-08 paper/paper.qmd
First pass of section Tensor Decompositions

The set of nonnegative numbers is denoted as $\mathbb{R}_+=\mathbb{R}_{\geq 0}=\left\{x\in \mathbb{R} \middle| x\geq 0\right\}$.

@@

We use a capital delta $\Delta$ to denote sets of vectors or higher order tensors where the slices or fibres along a specified dimension sum to $1$ i.e. generalized simplexes.

Usually, we use script letters ($\mathcal{A}, \mathcal{B}, \mathcal{C}, etc.$) for other sets.

@@

BlockTensorDecomposition.jl defines `identity_tensor`
### Operations

@@

## Common Decompositions
- Extensions of PCA/ICA/NMF to higher dimensions
- talk about the most popular Tucker, Tucker-n, CP
- other decompositions
  - high order SVD (see Kolda and Bader)
  - HOSVD (see Kolda, Shifted power method for computing tensor eigenpairs)

<!-- From my thesis proposal: -->

<!-- This section gives a brief tour of the most common tensor factorizations, heavily relying on the framework and notation given by Kolda and Bader [@kolda_TensorDecompositionsApplications_2009]. They are, however, presented in my own interpretation as matrix factorization generalizations and as different instances of the Tucker factorization.

One way we can extend matrix factorization to multidimensional signals is with a *Tucker-1* factorization. Here, each mixture is an $N$th order tensor $Y_i\in\mathbb{R}^{J_1\times \dots\times J_N}$ stored in the $1$-slices of an order-$(N+1)$ tensor $Y\in\mathbb{R}^{I\times J_1\times \dots\times J_N}$, $Y[i, :, \dots, :]=Y_i$, and similarly for $B$. Slice-wise, we have the equation

$$
Y_i = \sum_{r=1}^R A[i,r] B_r,
$$ {#eq-tucker-1-slices}

which is entry-wise

$$
Y[i, j_1, \dots, j_N] = \sum_{r=1}^R A[i,r] B[r,j_1, \dots, j_N].
$$ {#eq-tucker-1}

![Example of a rank $R=3$ Tucker-1 factorization coloured according to @eq-tucker-1-slices. Here we have $Y\in\mathbb{R}^{5\times J \times K}, A\in\mathbb{R}^{5\times 3}$, and $B\in\mathbb{R}^{3\times J \times K}$. The first horizontal slice of $Y$, $Y_1$, is equal to $A[1,1]B_1+A[1,2]B_2+A[1,3]B_3$.](figure/tucker_1_example.png){#fig-tucker-1-example}

Seeing @eq-tucker-1-slices alongside @fig-tucker-1-example makes it clear that the decomposition is performing a compression along the first axis of $Y$. This makes sense in the model where the mixtures are just linear combinations of sources with the same size and dimensions.

In the case where the sources can be further factored along other axes, specifically the first $n$ dimensions, we can use the more general *Tucker-$n$* factorization of an order $N$ tensor $Y$,

$$
\begin{split}
Y[i_1,\dots,i_n, & j_1, \dots, j_{N-n+1}] = \\ &\sum_{r_1=1}^{R_1} \dots \sum_{r_n=1}^{R_n} A_1[i_1,r_1]\cdots A_r[i_n,r_n] B[r_1,\dots,r_n, j_1, \dots, j_{N-n+1}].
\end{split}
$$ {#eq-tucker-n}

In the case where $n=N$ is the order of the tensor $Y$, we omit the $n$ and call this a *Tucker* factorization. So a Tucker factorization of an order $N$ tensor $Y$ is

$$
Y[i_1,\dots,i_N] = \sum_{r_1=1}^{R_1} \dots \sum_{r_n=1}^{R_n} A_1[i_1,r_1]\cdots A_r[i_n,r_N] B[r_1,\dots,r_N].
$$ {#eq-tucker}

All the Tucker decompositions can be written using the $n$-mode product $\times_n$ between a tensor $A$ and matrix $B$, defined as

$$
(A \times_n B)[i_1, \dots, i_{n-1},j,i_{n+1},\dots,N] = \sum_{i_{n}=1}^{I_n} A[i_1, \dots, i_{n-1},i_n,i-{n+1},\dots,N]B[i_n,j].
$$

They can also be written with an $n$ product shorthand, or with bracket notation. For example, the full tucker decomposition shown in @eq-tucker can be written as

$$
\begin{align*}
Y &= B\times_1 A_1 \times_2 \dots\times_N A_N = B \bigtimes_n A_n = \llbracket B; A_1, \dots, A_N \rrbracket \\
\text{or}\quad Y &= A_0\times_1 \dots\times_N A_N = A_0 \bigtimes_n A_n = \llbracket A_0; A_1, \dots, A_N \rrbracket.
\end{align*}
$$

To ease notation further, we sometimes consider the *core* tensor $B$ to be the zeroth factor $A_0=B$ so we can say $\{A_n\}_{n=0}^N$ are the factors of $Y$. Since it can be hard to visualize the product, consider the case of a Tucker factorization of a third order tensor $Y$:

$$
Y[i,j,k] = \sum_{r=1}^R\sum_{s=1}^S\sum_{t=1}^T A_1[i,r] A_2[j,s] A_3[k,t] B[r,s,t].
$$

We can picture this as multiplying the core factor $B$ with matrix factors $A_1, A_2, A_3$ on all sizes to construct $Y$ as shown in @fig-tucker.

@@

The Tucker-$n$ factorization of an order $N$ tensor can be thought of as a special case of the (full) Tucker factorization, where the matrix factors $A_{n+1}, \dots, A_{N}$ are identity matrices.

In the case where the core tensor $B\in\mathbb{R}^{R_1\times \dots\times R_N}$ of the Tucker factorization is cubic ($R_1=\dots=R_N$) and diagonal ($B[r_1,\dots,r_N]=\lambda_r$ if $r_1=\dots=r_N=r$ and $0$ otherwise), we have the **can**onical **decomp**osition/**para**llel **fac**tors model (CANDECOMP/PARAFAC or CP for short):

$$
Y[i_1,\dots,i_N] = \sum_{r=1}^{R} \lambda_r A_1[i_1,r]\cdots A_N[i_N,r].
$$

This can be written as

$$
Y = \llbracket \lambda; A_1, \dots, A_N\rrbracket.
$$

Sometimes the factor $\lambda_r$ is scaled so that the columns $A[:,r]$ are normalized $\lVert A[:,r]\rVert_2=1$ for each $r\in[R]$, or it is absorbed into the matrix factors and omitted from the decomposition. For the remainder of this text, I'll use CP decomposition to mean the latter case where $\lambda_r=1$ for all $r\in[R]$, as showing in @eq-cp:

$$
Y[i_1,\dots,i_N] = \sum_{r=1}^{R} A_1[i_1,r]\cdots A_N[i_N,r].
$$ {#eq-cp}

Other factorization models are used that combine aspects of CP and Tucker decomposition [@kolda_TensorDecompositionsApplications_2009], are specialized for order $3$ tensors [@qi_TripleDecompositionTensor_2020; @wu_manifold_2022], or provide alternate decomposition models entirely like tensor-trains [@oseledets_tensor-train_2011]. But the (full) Tucker, and its special cases Tucker-$n$, and CP decomposition are most commonly used extensions of the low-rank matrix factorization. These factorizations are summarized in @tbl-tensor-factorizations.

| Name       | Bracket Notation                             | $n$-mode Product                                         | Entry-wise   |
|------------|----------------------------------------------|----------------------------------------------------------|--------------|
| Tucker-$1$ | $\llbracket A_0; A_1 \rrbracket$             | $A_0\times_1 A_1$                                        | @eq-tucker-1 |
| Tucker-$n$ | $\llbracket A_0; A_1, \dots, A_n \rrbracket$ | $A_0\times_1 A_1 \times_2 \dots\times_n A_n$             | @eq-tucker-n |
| Tucker     | $\llbracket A_0; A_1, \dots, A_N \rrbracket$ | $A_0\times_1 A_1 \times_2 \dots\times_N A_N$             | @eq-tucker   |
| CP         | $\llbracket A_1, \dots, A_N \rrbracket$      | $\mathrm{id}_{R}\times_1 A_1 \times_2 \dots\times_N A_N$ | @eq-cp       |

: Summary of common tensor factorizations. Here, $N$ is the order of the factorized tensor. {#tbl-tensor-factorizations tbl-colwidths="[15,25,35,25]"}
 -->

@@

@@

>>>>

The set of real number is denoted as $\mathbb{R}$ and its restrictions to nonnegative numbers is denoted as $\mathbb{R}_+=\mathbb{R}_{\geq 0}=\left\{x\in \mathbb{R} \middle| x\geq 0\right\}$.

@@

We use a capital delta $\Delta$ to denote sets of vectors or higher order tensors where the slices or fibres along a specified dimension sum to $1$, i.e. generalized simplexes.

Usually, we use script letters ($\mathcal{A}, \mathcal{B}, \mathcal{C},$ etc.) for other sets.

@@

BlockTensorDecomposition.jl defines `identity_tensor(I, ndims)` to construct $\mathrm{id}_I$.

### Operations

@@

The $n$-slice dot product $\cdot_n$ between two tensors $A\in\mathbb{R}^{I_{1},\dots, I_{n-1},J,I_{n+1},\dots, I_{N}}$ and $B\in\mathbb{R}^{I_{1},\dots, I_{n-1},K,I_{n+1},\dots, I_{N}}$ returns a matrix $(A \cdot_n B)\in\mathbb{R}^{J\times K}$ with entries

$$
(A \cdot_n B)[j, k] = {\sum_{i_{1}\dots i_{n-1}i_{n+1}\dots i_{N}}} A[i_{1},\dots, i_{n-1},j,i_{n+1},\dots, i_{N}]B[i_{1},\dots, i_{n-1},k,i_{n+1},\dots, i_{N}].
$$

This product can also be thought of as taking the dot product between all pairs of $n$th order slices of $A$ and $B$: $(A \cdot_n B)[j, k] = A_j \cdot B_k$.

BlockTensorDecomposition.jl defines this operation with `slicewise_dot(A, B, n)`. In the special case where $A=B$, a more efficient method that only computes entries where $i \leq j$ is defined since $A \cdot_n A$ is a symmetric matrix.

The $n$-mode product $\times_n$ between a tensor $A\in\mathbb{R}^{I_1 \times \dots \times I_N}$ and matrix $B\in\mathbb{R}^{I_n\times J}$, returns a tensor $(A \times_n B)\in\mathbb{R}^{I_1 \times \dots \times I_{n-1} \times J \times I_{n+1} \times \dots \times I_N}$ with entries

$$
(A \times_n B)[i_1, \dots, i_{n-1},j,i_{n+1},\dots,i_N] = \sum_{i_{n}=1}^{I_n} A[i_1, \dots, i_{n-1},i_n,i_{n+1},\dots,i_N]B[i_n,j].
$$

BlockTensorDecomposition.jl defines this operation with `nmode_product(A, B, n)`.

```julia
function nmode_product(A::AbstractArray, B::AbstractMatrix, n::Integer)
    # convert the problem to the mode-1 product
    Aperm = swapdims(A, n)
    Cperm = Aperm ×₁ B
    return swapdims(Cperm, n) # swap back
end

function ×₁(A::AbstractArray, B::AbstractMatrix)
    # Turn the 1-mode product into matrix-matrix multiplication
    sizeA = size(A)
    Amat = reshape(A, sizeA[1], :)

    # Initialize the output tensor
    C = zeros(size(B)[1], sizeA[2:end]...)
    Cmat = reshape(C, size(B)[1], prod(sizeA[2:end]))

    # Perform matrix-matrix multiplication Cmat = B*Amat
    mul!(Cmat, B, Amat)

    return C # Output entries of Cmat in tensor form
end

function swapdims(A::AbstractArray, a::Integer, b::Integer=1)
    # Construct a permutation where a and b are swapped
    # e.g. [4, 2, 3, 1, 5, 6] when a=4 and b=1
    dims = collect(1:ndims(A))
    dims[a] = b; dims[b] = a
    return permutedims(A, dims)
end
```

The Frobenius norm of a tensor $A$ is the square root of its dot product with itself

$$
\left\lVert A \right\rVert_F = \sqrt{\left\langle A, A\right\rangle}.
$$

For vectors $v$, this is equivalent to the (Euclidean) 2-norm

$$
\left\lVert v \right\rVert_F = \left\lVert v \right\rVert_2 = \sqrt{\left\langle v, v\right\rangle}.
$$

For matrices $M$, the (Operator) 2-norm is defined as

$$
\left\lVert M \right\rVert_2 = \arg\,\max_{\left\lVert v \right\rVert_2 = 1} \left\lVert Mv \right\rVert_2 = \sigma_1(M)
$$

where $\sigma_1(M)$ is the largest singular value of $M$.

For tensors $T$, the (Operator) 2-norm needs to be defined in terms of how we treat them as function on other tensors. There is a canonical way to do this for vectors $x \to v^{\top}x$ and matrices $x \to Mx$, but not tensors. This is relevant to @sec-lipschitz-computation where the Lipschitz step-size is computed in terms of the Operator norm of the Hessian of our objective function.

## Common Decompositions {#sec-common-decompositions}
- Extensions of PCA/ICA/NMF to higher dimensions
- talk about the most popular Tucker, Tucker-n, CP
- other decompositions
  - high order SVD (see Kolda and Bader)
  - HOSVD (see Kolda, Shifted power method for computing tensor eigenpairs)

A tensor decomposition is a factorization of a tensor into multiple (usually smaller) tensors, that can be recombined into the original tensor. Computationally, we can think of a generic decomposition as storing factors $(A,B,C,...)$ and operations $(\times_a,\times_b,...)$ for combining them. This is what we do in BlockTensorDecomposition.jl.

```julia
struct GenericDecomposition{T, N} <: AbstractArray{T, N}
    factors::Tuple{Vararg{AbstractArray{T}}} # e.g. (A, B, C)
    contractions::Tuple{Vararg{Function}} # e.g. (×₁, ×₂)
end
# Y = A ×₁ B ×₂ C
array(G::GenericDecomposition) = multifoldl(contractions(G), factors(G))
```

The function `multifoldl` applies the given operations between each factor, from left to right.

```julia
function multifoldl(ops, args)
    @assert (length(ops) + 1) == length(args)
    x = args[begin]
    for (op, arg) in zip(ops, args[begin+1:end])
        x = op(x, arg)
    end
    return x
end
```

Different types of decompositions define different operations, and different "ranks" of the same decomposition specific the sizes of the factors used.

A commonly used family of decompositions can be derived from the Tucker decomposition.

::: {#def-tucker-decomposition}
A rank-$(R_1,\dots,R_N)$ Tucker decomposition of a tensor $Y\in\mathbb{R}^{I_1\times\dots\times I_N}$ produces $N$ matrices $A_n\in\mathbb{R}^{I_n\times R_n}$, $n\in[N]$, and core tensor $B\in\mathbb{R}^{R_1\times\dots\times R_N}$ such that

$$
Y[i_1,\dots,i_N] = \sum_{r_1=1}^{R_1} \dots \sum_{r_N=1}^{R_N} A_1[i_1,r_1]\cdots A_r[i_N,r_N] B[r_1,\dots,r_N]
$$ {#eq-tucker}

entry-wise. More compactly, this decomposition can be written using the $n$-mode product, or with double brackets

$$
Y = B\times_1 A_1 \times_2 \dots\times_N A_N = B \bigtimes_{n} A_n = \llbracket B; A_1, \dots, A_N \rrbracket.
$$

:::

Sometimes we write $A_0=B$ to ease notation, and suggest the "zeroth" factor of the tucker decomposition is the core tensor $B$. In the special case when $N=3$, we can visualize Tucker decomposition as multiplying the core tensor by matrices on all three sides as shown in @fig-tucker.

@@

Setting all the matrices of a Tucker decomposition to the identity matrix but the first gives the Tucker-$1$ decomposition.

::: {#def-tucker-1-decomposition}
A rank-$R$ Tucker-$1$ decomposition of a tensor $Y\in\mathbb{R}^{I_1\times\dots\times I_N}$ produces a matrix $A\in\mathbb{R}^{I_1\times R}$, and core tensor $B\in\mathbb{R}^{R\times I_2\times\dots\times I_N}$ such that

$$
Y[i_1,\dots,i_N] = \sum_{r=1}^{R} A[i_1,r] B[r, i_2,\dots,i_N]
$$ {#eq-tucker-1}

entry-wise or more compactly,

$$
Y = AB = B\times_1 A = \llbracket B; A \rrbracket.
$$

:::

Note we extend the usual definition of matrix-matrix multiplication

$$
(AB)[i,j] = \sum_{r=1}^R A[i,r] B[r, j]
$$

to tensors $B$ in the compact notation for Tucker-1 decomposition $Y=AB$.

More generally, any number of matrices can be set to the identity matrix giving the Tucker-$n$ decomposition.

::: {#def-tucker-n-decomposition}
A rank-$(R_1,\dots,R_n)$ Tucker-$n$ decomposition of a tensor $Y\in\mathbb{R}^{I_1\times\dots\times I_N}$ produces $n$ matrices $A_1,\dots,A_n$, and core tensor $B\in\mathbb{R}^{R_1\times\dots\times R_n \times I_{n+1}\times\dots\times I_N}$ such that

$$
Y[i_1,\dots,i_N] = \sum_{r_1=1}^{R_1} \dots \sum_{r_N=1}^{R_n} A_1[i_1,r_1]\cdots A_n[i_N,r_n] B[r_1,\dots,r_n,i_{n+1},\dots,i_{N}]
$$ {#eq-tucker-n}

entry-wise, or compactly written in the following three ways,
$$
\begin{align}
Y &= B\times_1 A_1 \times_2 \dots\times_n A_n \times_{n+1} \mathrm{id}_{I_{n+1}} \times_{n+2} \dots \times_{N} \mathrm{id}_{I_{N}}\\
Y &= B\times_1 A_1 \times_2 \dots\times_n A_n\\
Y &= \llbracket B; A_1, \dots, A_n \rrbracket.
\end{align}
$$

:::

Lastly, if we set the core tensor $B$ to the identity tensor $\mathrm{id}_{R}$, we obtain the **can**onical **decomp**osition/**para**llel **fac**tors model (CANDECOMP/PARAFAC or CP for short).

::: {#def-cp-decomposition}
A rank-$R$ CP decomposition of a tensor $Y\in\mathbb{R}^{I_1\times\dots\times I_N}$ produces $N$ matrices $A_n\in\mathbb{R}^{I_n\times R}$, such that

$$
Y[i_1,\dots,i_N] = \sum_{r=1}^{R} A_1[i_1,r]\cdots A_r[i_N,r]
$$ {#eq-cp}

entry-wise. More compactly, this decomposition can be written using the $n$-mode product, or with double brackets

$$
Y = \mathrm{id}_{R} \times_1 A_1 \times_2 \dots\times_N A_N = \mathrm{id}_{R} \bigtimes_{n} A_n = \llbracket A_1, \dots, A_N \rrbracket.
$$

:::

Note CP decomposition is sometimes referred to as Kruskal decomposition, and requires the core only be diagonal (and not necessarily identity) and the factors $A_n$ have normalized columns $\left\lVert A_n[:,r]\right\rVert_2 =1$.

Other factorization models are used that combine aspects of CP and Tucker decomposition [@kolda_TensorDecompositionsApplications_2009], are specialized for order $3$ tensors [@qi_TripleDecompositionTensor_2020; @wu_manifold_2022], or provide alternate decomposition models entirely like tensor-trains [@oseledets_tensor-train_2011]. But the (full) Tucker, and its special cases Tucker-$n$, and CP decomposition are most commonly used extensions of the low-rank matrix factorization to tensors. These factorizations are summarized in @tbl-tensor-factorizations.

| Name       | Bracket Notation                             | $n$-mode Product                                         | Entry-wise   |
|------------|----------------------------------------------|----------------------------------------------------------|--------------|
| Tucker     | $\llbracket A_0; A_1, \dots, A_N \rrbracket$ | $A_0\times_1 A_1 \times_2 \dots\times_N A_N$             | @eq-tucker   |
| Tucker-$1$ | $\llbracket A_0; A_1 \rrbracket$             | $A_0\times_1 A_1$                                        | @eq-tucker-1 |
| Tucker-$n$ | $\llbracket A_0; A_1, \dots, A_n \rrbracket$ | $A_0\times_1 A_1 \times_2 \dots\times_n A_n$             | @eq-tucker-n |
| CP         | $\llbracket A_1, \dots, A_N \rrbracket$      | $\mathrm{id}_{R}\times_1 A_1 \times_2 \dots\times_N A_N$ | @eq-cp       |

: Summary of common tensor factorizations. Here, $N$ is the order of the factorized tensor. {#tbl-tensor-factorizations tbl-colwidths="[15,25,35,25]"}

TODO add discussion on other decompositions
- high order SVD (see Kolda and Bader)
- HOSVD (see Kolda, Shifted power method for computing tensor eigenpairs)

Tensor decompositions are not nessisarily unique. It should be clear that scaling one factor by $x\neq 0$ and dividing another by $x$ yields the same original tensor. Furthermore, fibres and slices can be permuted without affecting the the original tensor. Up to these manipulations, for a fixed rank, there exist criteria that ensures their decompositions are unique [@kolda_TensorDecompositionsApplications_2009;@kruskal_three-way_1977;@bhaskara_uniqueness_2014].

### Representing Tucker Decompositions

There are implemented in BlockTensorDecomposition.jl and can be called, for a third order tensor, with `Tucker((B, A₁, A₂, A₃))`, `Tucker1((B, A₁))`, and `CPDecomposition((A₁, A₂, A₃))`. These Julia `structs` store the tensor in its factored form. If the recombined tensor or particular entries are requested, Julia dispatches on the type of decomposition and calls a particular method of `array` or `getindex`. The implementations for efficient array construction and index access are provided below.

```julia
array(T::Tucker) = multifoldl(tucker_contractions(ndims(T)), factors(T))
tucker_contractions(N) = Tuple((G, A) -> nmode_product(G, A, n) for n in 1:N)
```

TODO add getindex method for Tucker type

```julia
function array(T::Tucker1)
    B, A = factors(T)
    return B ×₁ A
end

function getindex(T::Tucker1, I::Vararg{Int})
    B, A = factors(T)
    i = I[1]
    J = I[begin+1:end] # J = (I₂, I₃, ..., I_N)
    return (@view A[i, :]) ⋅ view(B, :, J...)
end
```

```julia
array(CPD::CPDecomposition) =
  mapreduce(vector_outer, +, zip((eachcol.(factors(CPD)))...))

vector_outer(v) = reshape(kron(reverse(v)...),length.(v))

getindex(CPD::CPDecomposition, I::Vararg{Int}) =
  sum(reduce(.*, (@view f[i,:]) for (f,i) in zip(factors(CPD), I)))
```

@@

The rank of a matrix $Y\in\mathbb{R}^{I\times J}$ can be defined as the smallest $R\in\mathbb{Z}_+$ such that there exists an exact factorization $Y=AB$ for some $A\in\mathbb{R}^{I\times R}$ and $B\in\mathbb{R}^{R\times J}$.

Although this can be extended to higher order tensors, we must specify under which factorization model we are using. For example, the *CP-rank* $R$ of a tensor $Y$ is the smallest such $R$ that omits an exact CP decomposition of $Y$.

::: {#def-cp-rank}
The CP rank of a tensor $Y\in\mathbb{R}^{I_1\times \dots \times I_N}$ is the smallest $R$ such that there exist factors $A_n\in\mathbb{R}^{I_n \times R}$ and $Y = \llbracket A_1, \dots, A_N\rrbracket$,
$$
\text{rank}_\text{CP} (Y) = \min \left\{R \middle| \exists A_n\in\mathbb{R}^{I_n \times R},\, n\in[N] \quad \text{s.t.}\quad Y = \llbracket A_1, \dots, A_N\rrbracket \right\}
$$

:::

In a similar way, we can define the *Tucker-1-rank* $R$.

::: {#def-tucker-1-rank}
The Tucker-1 rank of a tensor $Y\in\mathbb{R}^{I_1\times \dots \times I_N}$ is the smallest $R$ such that there exist factors $A\in\mathbb{R}^{I_1 \times R}$ and $B\in\mathbb{R}^{R \times I_2 \times \dots \times I_N}$ where $Y = AB$

$$
\text{rank}_\text{Tucker-1} (Y) = \min \left\{R \middle| \exists A_n\in\mathbb{R}^{I_n \times R},B\in\mathbb{R}^{R \times I_2 \times \dots \times I_N}\, \quad \text{s.t.}\quad Y = AB\right\}
$$

:::

For the Tucker and Tucker-$n$ decompositions, we instead call a particular factorization **a** rank-$(R_1,\dots,R_N)$ Tucker factorization or **a** rank-$(R_1,\dots,R_n)$ Tucker-$n$ factorization, rather than **the** CP- or Tucker-$1$-rank of a tensor or **the** rank of a matrix.

One reason CP and Tucker-$1$ only need a single rank $R$ can be explained by considering the case when the order of the tensor $N=2$ (matrices). The two factorizations become equivalent and are equal to low-rank $R$ matrix factorization $Y=AB$. In fact, Tucker-$1$ is always equivalent to a low-rank matrix factorization, if you consider a flattening of the tensor to arrange the entries as a matrix.

The idea of tensor rank can be generalized further to constrained rank. These are the smallest rank $R$ such that the factors in the decomposition obey the given set of constraints.

For example, the nonnegative Tucker-1 rank is defined as
$$
\text{rank}_{\text{Tucker-1}}^+ (Y) = \min \left\{R \middle| \exists A_n\in\mathbb{R}_+^{I_n \times R},B\in\mathbb{R}_+^{R \times I_2 \times \dots \times I_N}\, \quad \text{s.t.}\quad Y = AB\right\}.
$$

More restrictive constraints increase the rank of the tensor since there is less freedom in selecting the factors.

@@

### Computing Gradients
- Use Auto diff generally
- But hand-crafted gradients and Lipschitz calculations *can* be faster (e.g. symmetrized slicewise dot product)

### Computing Lipschitz Step-sizes {#sec-lipschitz-computation}
